---
title: "Soybean Yield Response to Direct and Residual Phosphorus"
subtitle: "Linear Mixed Model Analysis and Visualization"
author: "Gustavo A. Roa"
date: "2026-04-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview


# Load Required Libraries

```{r}
library(tidyverse)    # Core tidyverse packages
library(readxl)       # read Excel files
library(lme4)         # mixed models
library(lmerTest)     # Kenward–Roger
library(emmeans)      # Estimated mean marginal means
library(multcomp)     # cld() for emmGrid
#library(writexl)      # write Excel files

```

# Data Preparation

## Load data

```{r}
# Import dataset
raw_data <- read_excel("soybean_P_response_dataset.xlsx", sheet = "data")

# Inspect structure
str(raw_data)

```

## Data setup

```{r}
# Convert relevant variables to factors
raw_data <- raw_data %>%
  mutate(across(c(loc_num, loc_name, trt_name), as.factor))

# Define treatment order (used in models and plots)
raw_data$trt_name <- factor(raw_data$trt_name,
  levels = c("0","34 R","67 R","101 R","134 R","34 D","67 D","101 D"))

```

## Analysis dataset

```{r}
# Standardize variable names for modeling
model_data <- raw_data %>%
  mutate(
    response = yield_mg_ha,
    trt      = trt_name,
    rep      = rep
  )

```

# Linear Mixed Model Analysis

## By-site models

For each site, the following linear mixed model was fitted:

$$
Y_{ijk} = \mu + \tau_i + b_j + \varepsilon_{ijk}
$$ where:

-   $Y_{ijk}$ is grain yield,

-   $\mu$ is the overall mean,

-   $\tau_i$ is the fixed effect of phosphorus treatment

-   $i$, $b_j \sim \mathcal{N}(0, \sigma_b^2)$ is the random effect of block (rep),

-   $\varepsilon_{ijk} \sim \mathcal{N}(0, \sigma^2)$ is the residual error.

This corresponds to: `lmer(response ~ trt + (1 | rep))`

```{r}
# Split dataset by site
data_list <- split(model_data, model_data$loc_num)

results_all_sites <- tibble()

for (i in data_list) {

  # Skip if insufficient variation
  if (nlevels(i$trt) < 2 || all(is.na(i$response))) next

  # Fit model
  model_site <- lmer(response ~ trt + (1 | rep),
                     data = i, REML = TRUE)

  # ANOVA
  anov <- anova(model_site, ddf = "Kenward-Roger")

  # Extract p-value
  p_value <- anov["trt", "Pr(>F)"]
  p_value_2 <- if (is.finite(p_value) && p_value < 0.001) "<0.001" else sprintf("%.3f", p_value)

  # Estimated means and grouping
  means <- emmeans(model_site, ~ trt, lmer.df = "kenward-roger")

  results <- cld(means, Letters = letters, adjust = "sidak") %>%
    as.data.frame() %>%
    mutate(
      letter      = trimws(.group),
      Sites       = as.character(i$loc_num[1]),
      Location    = as.character(i$loc_name[1]),
      p_value_num = p_value,
      P_Value     = p_value_2
    )

  # Store results
  results_all_sites <- bind_rows(results_all_sites, results)
}

```

## Across-site model

The combined analysis across sites was conducted using:

$$
Y_{ijkl} = \mu + \tau_i + s_k + b_{j(k)} + \varepsilon_{ijkl}
$$

where:

-   $Y_{ijkl}$ is grain yield,
-   $\mu$ is the overall mean,
-   $\tau_i$ is the fixed effect of phosphorus treatment $i$,
-   $s_k \sim \mathcal{N}(0, \sigma_s^2)$ is the random effect of site,
-   $b_{j(k)} \sim \mathcal{N}(0, \sigma_b^2)$ is the random effect of block nested within site,
-   $\varepsilon_{ijkl} \sim \mathcal{N}(0, \sigma^2)$ is the residual error.

This corresponds to:

`lmer(response ~ trt + (1 | loc_num/rep))`

```{r}
# Mixed model: treatment fixed, rep nested within location random
model_across <- lmer(response ~ trt + (1 | loc_num/rep),
                     data = model_data, REML = TRUE)

# ANOVA (Kenward-Roger correction)
anov_across <- anova(model_across, ddf = "Kenward-Roger")

# Extract p-value for treatment effect
p_value <- anov_across["trt", "Pr(>F)"]
p_value_2 <- if (is.finite(p_value) && p_value < 0.001) "<0.001" else sprintf("%.3f", p_value)

# Estimated marginal means and grouping
means_across <- emmeans(model_across, ~ trt, lmer.df = "kenward-roger")

results_across <- cld(means_across, Letters = letters, adjust = "sidak") %>%
  as.data.frame() %>%
  mutate(
    letter      = trimws(.group),
    Sites       = "Across",
    Location    = "All sites",
    p_value_num = p_value,
    P_Value     = p_value_2
  )

```

## Combine results

```{r}
# Merge across-site and site-specific results
results_final <- bind_rows(results_all_sites, results_across)

# Remove grouping letters when treatment effect is not significant
results_final$letter <- ifelse(
  is.na(results_final$p_value_num) | results_final$p_value_num > 0.05,
  "",
  results_final$letter
)

results_final
#write_xlsx(results_final, "./Anova results_final.xlsx")

```

# Visualization

## Prepare plotting dataset

```{r, fig.width=12, fig.height=8, warning=FALSE}

df_plot <- results_final %>%
  mutate(
    trt_label = factor(trt,
      levels = c("0","34 R","67 R","101 R","134 R","34 D","67 D","101 D")
    ),
    source = ifelse(grepl("D", trt),
                    "Direct P (soybean)",
                    "Residual P (corn)"),
    Sites = factor(Sites,
      levels = c("1","2","3","4","5","6","Across"))
  )

```

## Plot

```{r, fig.width=10, fig.height=14, warning=FALSE}

FigX <- ggplot(df_plot, aes(x = trt_label, y = emmean, color = source)) +
  
  geom_point(size = 4) +
  
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.2, linewidth = 0.8) +
  
  # Position letters relative to CI (consistent across panels)
  geom_text(aes(label = letter,
                y = upper.CL + 0.15 * (upper.CL - lower.CL)),
            color = "black", size = 5.5) +
  
  facet_wrap(~ Sites, scales = "free_y", ncol = 2) +
  
  scale_color_viridis_d(option = "cividis") +
  
  labs(x = expression(Phosphorus~rate~(kg~P[2]*O[5]~ha^-1)),
    y = expression(Grain~yield~(Mg~ha^-1)),
    color = "") +
  
  theme_classic(base_size = 22) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_rect(color = "black", linewidth = 1.5),
    legend.key = element_rect(color = "white"),
    legend.position = "top")

FigX

```

