---
title: "Soybean Yield Response to Direct and Residual Phosphorus"
subtitle: "Phosphorus Use Efficiency and Slope Comparison (LMM)"
author: "Gustavo A. Roa"
date: "2026-04-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview
This section evaluates the relative effects of residual phosphorus (P_corn) and direct phosphorus (P_soy) on soybean grain yield using a linear mixed model.

# Load Required Libraries
```{r}
library(tidyverse)    # Core tidyverse packages
library(readxl)       # read Excel files
library(lme4)         # mixed models
library(car)          # 

```

# Data Preparation
## Load data
```{r}
# Import dataset
raw_data <- read_excel("soybean_P_response_dataset.xlsx", sheet = "data")

# Inspect structure
#str(raw_data)

```

# Model 1: Yield response to P sources
$$
Y_{ij} = \mu + \beta_1 P_{\text{corn},ij} + \beta_2 P_{\text{soy},ij} + s_j + \varepsilon_{ij}
$$

where:

- $Y_{ij}$ is grain yield
- $\mu$ is the overall mean
- $\beta_1$ = yield increase per unit of residual P (efficiency)
- $\beta_2$ = yield increase per unit of direct P (efficiency)
- $s_j \sim \mathcal{N}(0, \sigma_s^2)$ = random effect of site
- $\varepsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$ = residual error

## Fit model
```{r}
# Linear mixed model:
model_eff <- lmer(yield_mg_ha ~ P_corn + P_soy + (1|loc_num), data = raw_data)

# Summary
summary(model_eff)
```

## Hypothesis test
Test whether residual and direct P effects are equal:
$$
H_0: \beta_1 = \beta_2
$$

```{r}
# Test equality of slopes (efficiency between sources)
car::linearHypothesis(model_eff, "P_corn = P_soy")

```




