---
title: "Soybean Yield Response to Direct and Residual Phosphorus"
subtitle: "Nonlinear Mixed Model and Critical P Estimation"
author: "Gustavo A. Roa"
date: "2026-04-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview
This section estimates the yield response and critical P levels for residual (corn) and direct (soybean) phosphorus using nonlinear mixed-effects models.

# Load Required Libraries
```{r}
library(tidyverse)    # Core tidyverse packages
library(readxl)       # read Excel files
library(nlme)         # Non Linear Mix Effects 

```

# Data preparation
## Load dataset
```{r}
# Import dataset
raw_data <- read_excel("soybean_P_response_dataset.xlsx", sheet = "data")

# Inspect structure
#str(raw_data)
```

## Split data by application pathway
```{r}
# Residual P (corn phase): no direct P applied
df_corn_raw <- raw_data %>% 
  filter(P_soy == 0)

# Direct P (soybean phase): no residual P applied
df_soy_raw  <- raw_data %>% 
  filter(P_corn == 0)

# Define a common P variable for modeling
df_corn_raw$P <- df_corn_raw$P_corn
df_soy_raw$P  <- df_soy_raw$P_soy

```


# Nonlinear mixed model (linear–plateau)
## Model definition
The model follows a linear–plateau response:
$$
Y = 
\begin{cases}
a + bP & \text{if } P < x_s \\
a + b x_s & \text{if } P \ge x_s
\end{cases}
$$

where:
- $\beta = initialslope (efficiency)$
- $\xs  = breakpoint (critical P)$ 

## Fit model: residual P (corn)
```{r}
nlmm_corn <- nlme(
  yield_mg_ha ~ ifelse(P < xs, a + b*P, a + b*xs),
  data = df_corn_raw,
  fixed = a + b + xs ~ 1,
  random = a ~ 1 | loc_num/rep,
  start = c(a = 3.0, b = 0.01, xs = 100),
  method = "REML"
)

summary(nlmm_corn)

```
### CI for residual P (corn)

```{r}
# Function to extract fixed effects
boot_fun <- function(fit) {
  fixef(fit)
}

# Bootstrap
set.seed(123)

sites <- unique(df_corn_raw$loc_num)
R <- 1000

boot_mat_R <- matrix(NA, nrow = R, ncol = 3)

for (r in 1:R) {
  
  # Sample sites with replacement
  sampled_sites <- sample(sites, size = length(sites), replace = TRUE)
  
  # Build bootstrap dataset (preserve full site structure)
  d <- dplyr::bind_rows(
    lapply(sampled_sites, function(s) {
      df_corn_raw %>% dplyr::filter(loc_num == s)
    })
  )
  
  # Fit model
  fit <- try(
    suppressWarnings(
      nlme(
        yield_mg_ha ~ ifelse(P < xs, a + b*P, a + b*xs),
        data = d,
        fixed = a + b + xs ~ 1,
        random = a ~ 1 | loc_num/rep,
        start = c(a = 3.0, b = 0.01, xs = 100),  # key difference for corn
        method = "REML",
        control = nlmeControl(msMaxIter = 100)
      )
    ),
    silent = TRUE
  )
  
  # Store results if valid
  if (!inherits(fit, "try-error")) {
    vals <- fixef(fit)
    if (all(is.finite(vals))) {
      boot_mat_R[r, ] <- vals
    }
  }
}

# Extract xs (3rd parameter)
xs_vals_R <- boot_mat_R[, 3]
xs_vals_R <- xs_vals_R[!is.na(xs_vals_R)]

# Confidence interval
xs_ci_R <- quantile(xs_vals_R, c(0.025, 0.975))

xs_ci_R

```

## Fit model: direct P (soybean)
```{r}
nlmm_soy <- nlme(
  yield_mg_ha ~ ifelse(P < xs, a + b*P, a + b*xs),
  data = df_soy_raw,
  fixed = a + b + xs ~ 1,
  random = a ~ 1 | loc_num/rep,
  start = c(a = 3.0, b = 0.01, xs = 40),
  method = "REML"
)

summary(nlmm_soy)

```

### CI for direct P (soybean)
```{r}
# Function to extract fixed effects
boot_fun <- function(fit) {
  fixef(fit)
}

# Bootstrap
set.seed(123)

sites <- unique(df_soy_raw$loc_num)
R <- 1000

boot_mat <- matrix(NA, nrow = R, ncol = 3)

for (r in 1:R) {
  
  # Sample sites with replacement
  sampled_sites <- sample(sites, size = length(sites), replace = TRUE)
  
  # Build bootstrap dataset (duplicate sites if sampled multiple times)
  d <- dplyr::bind_rows(
    lapply(sampled_sites, function(s) {
      df_soy_raw %>% dplyr::filter(loc_num == s)
    })
  )
  
  # Fit model
  fit <- try(
    suppressWarnings(
      nlme(
        yield_mg_ha ~ ifelse(P < xs, a + b*P, a + b*xs),
        data = d,
        fixed = a + b + xs ~ 1,
        random = a ~ 1 | loc_num/rep,
        start = c(a = 3.0, b = 0.01, xs = 40),
        method = "REML",
        control = nlmeControl(msMaxIter = 100)
      )
    ),
    silent = TRUE
  )
  
  # Store results if valid
  if (!inherits(fit, "try-error")) {
    vals <- fixef(fit)
    if (all(is.finite(vals))) {
      boot_mat[r, ] <- vals
    }
  }
}

# Extract xs (3rd parameter)
xs_vals_D <- boot_mat[, 3]
xs_vals_D <- xs_vals_D[!is.na(xs_vals_D)]

# Confidence interval
xs_ci_D <- quantile(xs_vals_D, c(0.025, 0.975))

xs_ci_D

```



## Model predictions

```{r}
# Generate prediction curves

# Prediction grids
new_corn <- data.frame(P = seq(1, 136, by = 1))
new_soy  <- data.frame(P = seq(1, 100, by = 1))

# Population-level predictions
new_corn$y_pred <- predict(nlmm_corn, newdata = new_corn, level = 0)
new_soy$y_pred  <- predict(nlmm_soy,  newdata = new_soy,  level = 0)

# Combine
new_corn$type <- "Corn"
new_soy$type  <- "Soy"

df_pred <- bind_rows(new_corn, new_soy)

```

## Observed summaries
```{r}
# Compute means and variability

# Residual P
df_obs_corn <- df_corn_raw %>%
  mutate(P = P_corn, type = "Corn") %>%
  group_by(type, P) %>%
  summarise(
    mean_y = mean(yield_mg_ha, na.rm = TRUE),
    sd = sd(yield_mg_ha, na.rm = TRUE),
    n = n(),
    se = sd / sqrt(n),
    .groups = "drop"
  ) %>%
  mutate(
    P_plot = P + 1.5,
    lower_sd = mean_y - sd,
    upper_sd = mean_y + sd
  )

# Direct P
df_obs_soy <- df_soy_raw %>%
  mutate(P = P_soy, type = "Soy") %>%
  group_by(type, P) %>%
  summarise(
    mean_y = mean(yield_mg_ha, na.rm = TRUE),
    sd = sd(yield_mg_ha, na.rm = TRUE),
    n = n(),
    se = sd / sqrt(n),
    .groups = "drop"
  ) %>%
  mutate(
    P_plot = P - 1.5,
    lower_sd = mean_y - sd,
    upper_sd = mean_y + sd
  )

df_obs <- bind_rows(df_obs_corn, df_obs_soy)

```

## Extract breakpoints
```{r}
xs_corn <- fixef(nlmm_corn)["xs"]
xs_soy  <- fixef(nlmm_soy)["xs"]

# Breakpoint data (so color maps correctly)
df_breaks <- data.frame(
  type = c("Corn", "Soy"),
  xs   = c(xs_corn, xs_soy),
  xmin = c(xs_ci_R[1], xs_ci_D[1]),
  xmax = c(xs_ci_R[2], xs_ci_D[2]),
  yend = c(3.6, 3.6)
)

# Create label
df_breaks$label <- sprintf(
  "%d (%d–%d)",
  round(df_breaks$xs),
  round(df_breaks$xmin),
  round(df_breaks$xmax)
)

```

# Visualization

```{r, fig.width=12, fig.height=8, warning=FALSE}

FigX <- ggplot() +
  
  # Model predictions
  geom_line(data = df_pred,
            aes(P, y_pred, color = type),
            linewidth = 2) +
  
  # Observed means
  geom_point(data = df_obs,
             aes(P_plot, mean_y, color = type),
             size = 5) +
  
  # Variability (SD)
  geom_errorbar(data = df_obs,
                aes(P_plot, ymin = lower_sd, ymax = upper_sd, color = type),
                width = 2,
                linewidth = 1) +
  
  # Breakpoints
  geom_segment(data = df_breaks,
               aes(x = xs, xend = xs, y = 0, yend = yend, color = type),
               linetype = "dashed",
               linewidth = 1.5) +
  
  geom_text(data = df_breaks,
            aes(x = xs + 2, y = 1.2, label = label),
            color = "black",
            size = 6,
            hjust = 0,
            show.legend = FALSE) +
  
  scale_y_continuous(limits = c(0, 5), breaks = seq(0, 5, 1), expand = c(0,0)) +
  scale_x_continuous(limits = c(0, 140), breaks = c(0, 34, 67, 101, 134)) +
  
  scale_color_viridis_d(
    option = "cividis",
    labels = c("Corn" = "Residual P (corn)",
               "Soy"  = "Direct P (soybean)")
  ) +
  
  labs(
    x = expression(Phosphorus~rate~(kg~P[2]*O[5]~ha^-1)),
    y = expression(Grain~yield~(Mg~ha^-1)),
    color = ""
  ) +
  
  theme_classic(base_size = 22) +
  theme(
    axis.text = element_text(color = "black"),
    panel.background = element_rect(color = "black", linewidth = 2),
    legend.key = element_rect(color = "white"),
    legend.position = "top"
  )

FigX 

```

